---
title: "Lab 04 — Sampling from Time Series"
author: "EAES 480 Dylan Enwia — Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '2 16 2026')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: '3'
editor_options:
  chunk_output_type: inline
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

## What is AmeriFlux?

**AmeriFlux** is a network of **eddy covariance** flux-tower sites that measure exchanges of **carbon (CO₂), water, and energy** between ecosystems and the atmosphere, with standardized data products shared for research and education.

In this lab, you will use a simplified AmeriFlux-style dataset from the **US-AMS site at Argonne National Laboratory (near Chicago)**. The measurements are at **30-minute resolution** over **2023**, and show strong **seasonality** and **day–night cycles**.

**Key idea for this lab:** treat the full 2023 time series as the **population**, then practice sampling strategies to estimate population parameters.

References for context:
- AmeriFlux overview: https://ameriflux.lbl.gov/about/about-ameriflux/
- US-AMS site page: https://ameriflux.lbl.gov/sites/siteinfo/US-AMS

---

# Learning goals

By the end of this lab, you should be able to:

- Define a **population** and a **sample** for an EAES time-series dataset
- Compute population **parameters** (mean, SD) and compare to sample **estimates**
- Visualize distributions and identify **latent grouping variables** (month, day/night)
- Implement **simple random sampling** and **stratified sampling**
- Use `set.seed()` to make sampling reproducible

---

# Data

## Load and inspect

```{r load_data, echo=TRUE, eval=FALSE}
df <- read_csv("data/us-ams-simple.csv",
               na = c("-9999")) %>%
  clean_names() %>% filter(year_local == 2023)

glimpse(df)
```

**CHECK:** You should see columns like `year_local`, `doy`, `daytime`, and flux/biomet variables (e.g., `gpp`, `fc`, `le`, `ta`).

---

## Create a date and month column from DOY

This dataset uses **Year + Day-of-Year (DOY)**. Month must be derived from a calendar date.

```{r derive_time, echo=TRUE, eval=FALSE}
df <- df %>%
  filter(year_local == 2023) %>%
  filter(daytime != -9999) %>% 
  filter(doy != -9999) %>%
  mutate(
    # TODO: create a Date column from year_local and doy
    # HINT: Jan 1 is DOY = 1, so use (doy - 1) with origin = "YYYY-01-01"
    date = as.Date(doy - 1, origin = paste0(year_local, "-01-01")),

    # TODO: create a month column (numeric 1–12 or labeled months)
    month = month(date, label = TRUE, abbr = TRUE),

    # TODO: make a day/night label using daytime (0/1)
    day_night = if_else(daytime == 1, "Day", "Night")
  )
count(df, month)
count(df, day_night)
```

---

# Choose a response variable

You will analyze **one response variable** throughout the lab. This could be a CO₂ flux metric or a meteorological variable.

Examples you can choose from (depending on what you see in the dataset):
- CO₂ / carbon: `gpp`, `reco`, `fc`
- Energy: `le`
- Meteorology: `ta`, `ts`, `swc`

```{r choose_response, echo=TRUE, eval=FALSE}
# TODO: choose ONE response variable (a column name as a string)
response_var <- "ta"   # replace "gpp" with your choice, e.g. "fc" or "le" or "ta"

# CHECK: print a quick summary
df %>% summarise(
  n = n(),
  n_missing = sum(is.na(df[[response_var]])),
  mean = mean(df[[response_var]], na.rm = TRUE),
  sd = sd(df[[response_var]], na.rm = TRUE)
)
```

**Prompt (2–3 sentences):** Why did you choose this response variable? What do you expect its seasonality/day–night pattern to be?

> *I chose air temperature as my response variable because I think it controls many flux variables and is easy to interpret. I would expect its seasonality/day-night pattern to be having higher daytime temperatures and cooler nighttime temperatures and a seasonal cycle with a very cold winter and hot summer.*

---

# Section 1 — Data dictionary (conceptual)

Students will populate the data dictionary using:
https://ameriflux.lbl.gov/data/aboutdata/data-variables/

Fill in at least **5 variables** from this dataset:

| Variable | Units | Description | Expected sign/seasonality? |
|----------|-------|-------------|----------------------------|
| PA | kPA |Atmospheric pressure|No expected sign/seasonality|
| TA | deg C |Air Temperature | High seasonality (cold in winter and warm in summer)|
| D_SNOW | cm | Snow depth  |High in winter seasons|
| SLE |	W m-2 | Latent heat (LE) storage flux |Seasonal cycle (higher in growing season)|
| SAP_FLOW |mmolH2O m-2 s-1 |Sap Flow|Seasonal (higher in growing season)|

---

# Section 2 — Visualizing the population

Remember: for this lab, the **population** is the entire 2023 half-hourly time series.

## 2.1 Time series view

```{r plot_time_series, echo=TRUE, eval=FALSE}
# GOAL: Visualize seasonality over the year.
# TODO: pick a y aesthetic using response_var.

df %>%
  ggplot(aes(x = date, y = df[[response_var]])) +
  geom_line(alpha = 0.25) +
  theme_classic(base_size = 18) +
  labs(
    x = NULL,
    y = response_var,
    title = "Population time series (2023)"
  )
```

**Prompt (2–3 sentences):** What major patterns do you see? (Seasonal cycle? Daily cycle? Outliers?)

> *Major patterns I see is that temperatures are lowest in the winter and highest in the summer. We can see for the daily cycle that values rise during the day and fall at night. Outliers I see for example would be any sharp spikes or gaps that could reflect from remaining missing data points.*

---

## 2.2 Population distribution (histogram + density)

```{r pop_distribution, echo=TRUE, eval=FALSE}
# GOAL: See the overall distribution of the population.
# TODO: choose an appropriate number of bins (start with ~50).

ggplot(df, aes(x = df[[response_var]])) +
  geom_histogram(bins = 50, alpha = 0.7) +
  theme_classic(base_size = 18) +
  labs(x = response_var, y = "Count", title = "Population distribution (histogram)")

ggplot(df, aes(x = df[[response_var]])) +
  geom_density(alpha = 0.7) +
  theme_classic(base_size = 18) +
  labs(x = response_var, y = "Density", title = "Population distribution (density)")
```

**Prompt:** Describe shape (skew, modality), center, and spread.

> *The distribution shows its bimodal so we can see a separation between winter and summer conditions. The center shows to be around (10-15 Celsius) for the annual average. And also we can see a large spread which shows a high seasonal variability due to higher cold and hot temperature values. *

---

## 2.3 Do latent groups explain variability? (month, day/night)

### By month

```{r pop_by_month, echo=TRUE, eval=FALSE}
# TODO: make month appear in a sensible order (it already is an ordered factor if label=TRUE)
df %>%
  ggplot(aes(x = month, y = df[[response_var]])) +
  geom_boxplot(outlier.alpha = 0.25) +
  theme_classic(base_size = 18) +
  labs(x = NULL, y = response_var, title = "Population by month")
```

### By day/night

```{r pop_by_daynight, echo=TRUE, eval=FALSE}
df %>%
  ggplot(aes(x = day_night, y = df[[response_var]])) +
  geom_boxplot(outlier.alpha = 0.25) +
  theme_classic(base_size = 18) +
  labs(x = NULL, y = response_var, title = "Population by day vs night")
```

**Prompt (3–4 sentences):** Which grouping variable (month or day/night) seems to explain more variability in your response? Why?

> *I would say the month grouping variable explains more variability in air temperatures because seasonal temperature range between winter to summer is higher than a day to night change. The monthly box plots show a strong seasonal variability when compared to the day/night box plots. We can see this by increasing values starting at 0 Celsius in winter and going to 20 Celsius in the summer then decreasing again into the fall and winter while in the day/night box plot we are seeing an overlap between the day and night groups.*

---

# Section 3 — Population parameters (truth)

Compute the population mean and SD for your chosen response variable.

```{r population_params, echo=TRUE, eval=FALSE}
pop_mean <- mean(df[[response_var]], na.rm = TRUE)
pop_sd   <- sd(df[[response_var]], na.rm = TRUE)

tibble(
  response_var = response_var,
  population_mean = pop_mean,
  population_sd = pop_sd
)
```

---

# Section 4 — Simple random sampling (SRS)

## 4.1 One random sample

```{r one_sample, echo=TRUE, eval=FALSE}
set.seed(480)

# TODO: choose a sample size (e.g., 200, 500, 1000)
n_samp <- 1000

samp <- df %>%
  slice_sample(n = n_samp)

samp_mean <- mean(samp[[response_var]], na.rm = TRUE)
samp_sd   <- sd(samp[[response_var]], na.rm = TRUE)

tibble(
  n_samp = n_samp,
  sample_mean = samp_mean,
  sample_sd = samp_sd,
  pop_mean = pop_mean,
  pop_sd = pop_sd
)
```

**Prompt (2–3 sentences):** How close is your one-sample estimate to the population mean/SD? Is the difference surprising?

> *The one-sample estimate and the population mean/sd are very close with a small difference. The difference does not surprise me at all because we have a large sample size which should closely approximate the population mean/sd.*

---

## 4.2 Sampling variability: many samples → many means

```{r sampling_distribution, echo=TRUE, eval=FALSE}
set.seed(480)

reps <- 500   # TODO: choose number of replicates (e.g., 500 or 1000)

means <- replicate(
  reps,
  df %>%
    slice_sample(n = n_samp) %>%
    summarise(m = mean(ta, na.rm = TRUE)) %>%
    pull(m)
)

ggplot(tibble(mean_est = means), aes(mean_est)) +
  geom_histogram(bins = 40, alpha = 0.8) +
  geom_vline(xintercept = pop_mean, linetype = "dashed", linewidth = 1.1) +
  theme_classic(base_size = 18) +
  labs(
    x = paste0("Sample mean of ", response_var),
    y = "Count",
    title = "Sampling distribution of the mean (SRS)",
    subtitle = "Dashed line = population mean"
  )
```

**Prompt (2–3 sentences):** Is the sampling distribution centered on the population mean? What happens if you increase `n_samp`?

> *Yes the sampling distribution is centered on the population mean. If you increase n_samp the sampling distribution would start to become narrower or smaller due to a reduction in standard error.*

---

# Section 5 — Stratified sampling

Here you’ll test whether stratification helps when the population has structure.

## 5.1 Stratify by month

```{r strat_by_month, echo=TRUE, eval=FALSE}
set.seed(480)

# GOAL: sample within each month to ensure seasonal representation.
# TODO: choose n_per_month so total sample size is reasonable (e.g., 12 * 20 = 240)
n_per_month <- 1200

samp_strat <- df %>%
  group_by(month) %>%
  slice_sample(n = n_per_month) %>%
  ungroup()

strat_mean <- mean(samp_strat[[response_var]], na.rm = TRUE)
strat_sd   <- sd(samp_strat[[response_var]], na.rm = TRUE)

tibble(
  n_per_month = n_per_month,
  total_n = nrow(samp_strat),
  strat_mean = strat_mean,
  strat_sd = strat_sd,
  pop_mean = pop_mean,
  pop_sd = pop_sd
)
```

---

## 5.2 Compare strategies (SRS vs stratified)

```{r compare_sampling, echo=TRUE, eval=FALSE}
tibble(
  strategy = c("Population", "SRS", "Stratified by month"),
  mean = c(pop_mean, samp_mean, strat_mean),
  sd   = c(pop_sd,   samp_sd,   strat_sd)
)
```

**Prompt (3–4 sentences):** Which strategy better approximated the population mean and SD for your response variable? Why might stratification help (or not) here?

> *I would say the stratified by month strategy would better approximate the population mean and SD for air temperature. I say this because its pretty much almost identical to the population mean and SD. The SRS mean on the other hand is farther from the population values. Stratification can help here because we saw temperature have a strong seasonal variability and when sampling evenly within each month we can ensure that all seasons can be represented properly.*

---

# Section 6 — Conceptual reflection

Answer in **4–6 sentences**:

- Why does seasonality matter for sampling?
- What happens if sampling ignores latent grouping variables?
- In EAES field studies, when is stratification essential?
- What is one trade-off of stratified sampling?

> *Seasonality matters for sampling because the seasonlity mean and variability changes through the year so sampling has to represent every season to measure it properly. If sampling ignores latent grouping variables like for example month or day/night it can over estimate certain times like winter nights or summer afternoons and from this we can result in biased estimations in the mean and SD. Straftification is essential in EAES field studies because conditions often change for example we can look at a wet vs dry season, land cover types, growing vs non-growing season, and etc, so we need each variable to be properly represented. The one trade-off stratified sampling is that it can be very complex and costly in a study.*

---

# Part II — Sampling designs extensions (graded practice)

In Part I you treated the full 2023 half-hourly record as a **population**, then compared **simple random sampling** (SRS) vs **stratified sampling by month**.

Now you will practice additional **sampling designs** discussed in lecture:

- **Systematic** sampling (regular interval)
- **Cluster** sampling (sample groups, then measure everything in them)
- **Quasi-continuous** sampling (regular time series subsampling)
- **Blocked** designs (preview only — think “blocks as structure”)

All exercises below should run using the same objects from Part I:
- `df` (with `date`, `month`, `day_night`)
- `response_var`
- `pop_mean`, `pop_sd`
- your SRS sample `samp` and stratified sample `samp_strat` (if you created them)

> **Tip:** If you renamed objects in Part I, update the code below to match your names.

---

## Exercise 1 — Systematic sampling (every k-th observation)

**Idea:** sample at a fixed interval (e.g., every 48th record ≈ daily at 30-min resolution).

**Risk:** if the variable has strong cycles aligned with the interval, systematic sampling can be biased.

```{r systematic_sampling, echo=TRUE, eval=FALSE}
# GOAL: Create a systematic sample and compare mean/sd to population.
# TODO: choose an interval k (try 48, 24, 96).
k <- 48

sys_samp <- df %>%
  # TODO: keep only non-missing response values
  filter(!is.na(.data[[response_var]])) %>%
  slice(seq(1, n(), by = k))

sys_mean <- mean(sys_samp[[response_var]], na.rm = TRUE)
sys_sd   <- sd(sys_samp[[response_var]], na.rm = TRUE)

tibble(
  strategy = c("Population", "Systematic"),
  mean = c(pop_mean, sys_mean),
  sd   = c(pop_sd,   sys_sd),
  n    = c(nrow(df), nrow(sys_samp))
)
```

**Prompt (3–4 sentences):** Did systematic sampling approximate the population mean/SD better or worse than SRS?  
What cycle (daily or seasonal) might be interacting with your chosen `k`?

> *The systematic sampling approximated very well for the population mean. For the SD it was worse it was higher in the systematic sampling so it didn't approximate the population SD as well as the SRS. I think the daily cycle is interacting with my chosen `k`. *

---

## Exercise 2 — Cluster sampling (sample days, take all points within those days)

**Definition:** choose a set of clusters (here, **days**) at random, then include **all observations** in the chosen clusters.

This mimics EAES logistics: you may only be able to sample on certain days.

```{r cluster_sampling_days, echo=TRUE, eval=FALSE}
# GOAL: Sample whole days as clusters, then estimate mean/sd.
# TODO: choose number of days to sample.
set.seed(480)

n_days <- 30

days <- df %>%
  distinct(date) %>%
  drop_na(date) %>%
  pull(date)

chosen_days <- sample(days, size = n_days, replace = FALSE)

cluster_samp <- df %>%
  filter(date %in% chosen_days) %>%
  filter(!is.na(.data[[response_var]]))

clust_mean <- mean(cluster_samp[[response_var]], na.rm = TRUE)
clust_sd   <- sd(cluster_samp[[response_var]], na.rm = TRUE)

tibble(
  strategy = c("Population", "Cluster (days)"),
  mean = c(pop_mean, clust_mean),
  sd   = c(pop_sd,   clust_sd),
  n    = c(nrow(df), nrow(cluster_samp))
)
```

**Prompt (3–4 sentences):** Why might cluster sampling have **higher variance** than SRS for the same number of measurements?  
What feature of time series data (hint: autocorrelation) is relevant here?

> *Cluster sampling can have higher variance than SRS because observations within the same day are similar due to autocorrelation. Even though since the sample size is large (n = 1332) much of that information is inessential because the values from each selected day are correlated. This explains why cluster sampling may perform worse than SRS for estimating population mean/SD.*

---

## Exercise 3 — Quasi-continuous sampling (fixed schedule time series)

**Definition:** sample regularly over time to create a *subsampled time series*.

This mimics continuous instrumentation that logs at a lower frequency (e.g., hourly instead of 30-min).

```{r quasi_continuous, echo=TRUE, eval=FALSE}
# GOAL: Create a regular subsampled time series.
# TODO: choose a step size (every 2 records = hourly; every 4 = 2-hourly).
step <- 2

qc_samp <- df %>%
  filter(!is.na(.data[[response_var]])) %>%
  slice(seq(1, n(), by = step))

qc_mean <- mean(qc_samp[[response_var]], na.rm = TRUE)
qc_sd   <- sd(qc_samp[[response_var]], na.rm = TRUE)

tibble(
  strategy = c("Population", "Quasi-continuous"),
  mean = c(pop_mean, qc_mean),
  sd   = c(pop_sd,   qc_sd),
  n    = c(nrow(df), nrow(qc_samp))
)
```

### Visual check: does the subsampled series preserve structure?

```{r qc_plot, echo=TRUE, eval=FALSE}
# TODO: plot BOTH population and qc_samp time series (thin lines) for a short window
# HINT: filter to one month to avoid overplotting.

df %>%
  filter(month == "May") %>%   # e.g., "Jul" if month is labeled
  ggplot(aes(x = date, y = .data[[response_var]])) +
  geom_line(alpha = 0.25) +
  theme_classic(base_size = 18) +
  labs(title = "Population time series (subset)", x = NULL, y = response_var)

qc_samp %>%
  filter(month == "May") %>%
  ggplot(aes(x = date, y = .data[[response_var]])) +
  geom_line(alpha = 0.6) +
  theme_classic(base_size = 18) +
  labs(title = "Quasi-continuous sample time series (subset)", x = NULL, y = response_var)
```

**Prompt (3–4 sentences):** Does quasi-continuous sampling preserve the **seasonal** pattern? The **daily** pattern?  
In an EAES context, when is quasi-continuous sampling preferable to sparse discrete sampling?

> *The quasi-continuous sampling preserves daily pattern. We can see the quasi-continuous sampling mean/SD and the population mean/SD are nearly identical. In an EAES context, quasi-continuous sampling is preferable when you want to gather a continuous record of data like daily cycles for example. I think it is more informative than sparse discrete sampling because its keeps the timing more efficiently of a process of data that you want to study.*

---

## Exercise 4 — Compare all strategies in one table (and interpret)

```{r compare_all, echo=TRUE, eval=FALSE}
# GOAL: Assemble a comparison table of mean/sd error for each strategy.
# NOTE: This assumes you created objects: samp, samp_strat, sys_samp, cluster_samp, qc_samp.
# If your object names differ, update them here.

summ_stats <- function(dat, label) {
  tibble(
    strategy = label,
    n = nrow(dat),
    mean = mean(dat[[response_var]], na.rm = TRUE),
    sd   = sd(dat[[response_var]], na.rm = TRUE)
  )
}

bind_rows(
  tibble(strategy = "Population", n = nrow(df), mean = pop_mean, sd = pop_sd),
  summ_stats(samp, "SRS"),
  summ_stats(samp_strat, "Stratified (month)"),
  summ_stats(sys_samp, "Systematic"),
  summ_stats(cluster_samp, "Cluster (days)"),
  summ_stats(qc_samp, "Quasi-continuous")
) %>%
  mutate(
    mean_error = mean - pop_mean,
    sd_error   = sd - pop_sd
  )
```

**Prompt (5–6 sentences, graded):** Which strategy gave the best estimate of the population mean? Of the population SD?  
Explain *why* in terms of (i) seasonal/diurnal structure and (ii) dependence/autocorrelation.  
If you were designing a real EAES study with limited field days, what hybrid strategy would you propose (e.g., stratified + clustered)?

> *The strategy that best estimated the population mean was the Stratified (month). The strategy that best estimated the population SD was also the Stratified (month). SRS performed reasonably well but showed slightly more deviation, while Systematic sampling overestimated variability and Cluster sampling showed the largest deviation in the mean. These differences showed a strong seasonal/diurnal structure of temperature, as well as autocorrelation in the time series data due to cluster sampling performing worse because observations collected within the same day were highly correlated which means they are dependent. If I was designing a real EAES study with limited field days, a strong hybrid approach would be stratified and clustered strategies. I would select days within a season or month to ensure seasonal coverage and sample thoroughly within those selected days to accurately capture daily variability.*

---

## Blocked designs (preview only — do not implement yet)

A **blocked** design means sampling within blocks (time blocks like months, or spatial blocks like sites), then later treating block as structure in the model (often as a random effect).

You already approximated blocking by stratifying over `month`. Later, we will return to this idea when we fit models that include block structure explicitly.

# Submission + self-check

## Before you knit (Run All)
- [ ] Setup chunk runs without errors
- [ ] Data join happened (df_raw has both canopy + climate columns)
- [ ] All chunks run top-to-bottom
- [ ] No objects created only in the Console
- [ ] Interpretation answers are complete sentences

## After you knit
- [ ] Figures appear in the output
- [ ] Tables render and are readable
- [ ] Your selected x and y are clearly stated in the document

**Save, commit, push to Github.**
